---
output:
  pdf_document:
          fig_caption: yes
          number_sections: yes
bibliography: tfm.bib
csl: american-statistical-association.csl
urlcolor: black
linkcolor: black
fontsize: 11pt
geometry: margin = 1.2in
header-includes:
- \usepackage{amsmath}
- \usepackage{placeins}
- \usepackage{fancyhdr}
- \usepackage{setspace}
- \usepackage{chngcntr}
- \usepackage{microtype}
- \onehalfspacing
- \counterwithin{figure}{section}
- \counterwithin{table}{section}
---

```{r global_options, include=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)
library(ggridges)
library(mice)
library(scales)
library(lme4)
library(geepack)
opts_chunk$set(fig.path = 'figures/', echo = FALSE, warning = FALSE, message = FALSE,
               fig.align = "center")
```

\pagenumbering{gobble}

```{r child = "title_page.Rmd"}
```

\newpage

```{r child = "license.Rmd"}
```

\newpage
\pagestyle{fancy}

\fancyhead[LE,RO]{}
\fancyhead[LO,RE]{}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0pt}

\pagenumbering{roman}

\fancyhead[CO,CE]{Abstract}
\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}

```{r child = "abstract.Rmd"}
```


\newpage
\fancyhead[CO,CE]{Table of Contents}
\setcounter{tocdepth}{2}
\tableofcontents

\newpage
\pagenumbering{arabic}

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Introduction}

```{r child = "introduction.Rmd"}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Clustered data}


```{r child = "clustered_data.Rmd"}

```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Linear mixed models}

```{r child = "lmm.Rmd"}

```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Generalized estimating equations}

```{r child = "gee.Rmd"}

```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Missing data}

```{r child = "missing.Rmd"}

```


\FloatBarrier
\newpage
\fancyhead[CO,CE]{LAKE study}

```{r child = "lake_data.Rmd"}

```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Microarray study}

```{r child = "microarray_study.Rmd"}

```


\FloatBarrier
\newpage
\fancyhead[CO,CE]{Conclusion}

```{r child = "discussion.Rmd"}

```

\FloatBarrier

\newpage
\fancyhead[CO,CE]{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures

\newpage
\fancyhead[CO,CE]{List of Tables}
\addcontentsline{toc}{section}{List of Tables}
\listoftables


\FloatBarrier
\newpage
\fancyhead[CO,CE]{References}

# References